Variable Selection in Convex Function Estimation

### 2014-04-24 tex/scam/optimization.tex
experiment code

The sparse convex additive model optimization in (\ref{np}) is a quadratic program (QP) with
$O(np)$ variables and $O(np)$ constraints. 
Directly applying a QP solver for $\bds{h},
\bds{\beta}$ would be computationally expensive for relatively large
$n$ and $p$. However, notice that variables in different feature
dimensions are only coupled in the term
$(Y_{i}-\sum_{k=1}^{p}h_{ki})^{2}$. Hence, we can apply the block
coordinate descent method, where in each step we solve the following
QP subproblem for $\{\bds{h}_{k\cdot},\bds{\beta}_{k\cdot}\}$ with the
other variables fixed. We present the one-block optimization in matrix notation:
\begin{align*}
\min_{ \bds{f}_k, \bds{\beta}_k, \gamma_k}& \frac{1}{2n} \| \bds{r}_{\mh k} - \bds{f}_k \|_2^2 
     + \lambda \gamma_k \\
 \textrm{s.t. } & P \bds{f}_k = \diag(P \bds{x}_k)  \bds{\beta}_k \\
   & D \bds{\beta}_k \leq 0 ,
   \quad -\gamma_k \mathbf{1}_n \leq \bds{f}_k \leq \gamma_k \mathbf{1}_n,
   \quad \mathbf{1}_n^\tran \bds{f}_k = 0
\end{align*}

@@

\begin{equation}
\label{eqn:opt_1d}
\begin{split}
       \min_{\bds{h}_{k\cdot},\bds{\beta}_{k\cdot},\gamma_{k}} &
             \ \frac{1}{2n}\sum_{i=1}^{n}\Bigl((Y_{i}-\bar{Y}
                -\sum_{r\neq{k}}f_{ri})-f_{ki}\Bigr)^{2} 
                      + \lambda\gamma_{k} \\
        \textrm{such that} & \ f_{k(i+1)} = f_{k(i)} + \beta_{k(i)}(x_{k(i+1)}-x_{k(i)}),\\
        &\ \beta_{k(i+1)} \geq \beta_{k(i)}, \ -\gamma_{k}\leq f_{ki}\leq\gamma_{k}\\
        &\  \sum_{i=1}^{n}f_{ki}=0, \ (\forall i).
\end{split}
\end{equation}
The extra variable $\gamma_{k}$ is introduced to deal with the $\ell_{\infty}$ norm. This QP subproblem involves $O(n)$ variables, $O(n)$ constraints and a sparse structure, 
which can be solved efficiently using optimization packages (e.g., MOSEK: \verb+http://www.mosek.com/+).  We cycle through all feature dimensions ($k$) from $1$ to $p$ multiple times until convergence.
Empirically, we observe that the algorithm converges in only a few cycles. We also implemented an ADMM solver for (\ref{np}), but found that it is not as efficient as this QP solver.

After optimization, the function estimator for any input data $\bds{x}_j$ is, according to (\ref{hyper}),
\begin{equation}
\begin{split}
\nonumber
      f(\bds{x}_j) & = \sum_{k=1}^{p}f_k(x_{kj})+\mu \\ 
    & = \sum_{k=1}^{p}\max_{i} \{f_{ki}+\beta_{ki}(x_{kj}-x_{ki})\} +
      \mu.
\end{split}
\end{equation}

The univariate concave function estimation is a straightforward modification of optimization~\ref{eqn:opt_1d}. We need only modify one linear inequality constraint to enforce that the subgradients must be non-increasing: $\beta_{k(i+1)} \leq \beta_{k(i)}$.

@@

It is easy to verify that $\beta_{k(i)} = \sum_{j \leq i} d_{k(j)}$ and 
\begin{align*}
f_k(x_{k(i)}) = & f_k(x_{k(i-1)}) + \beta_{k(i-1)}(x_{k(i)} - x_{k(i-1)}) \\
 =& f_k(x_{k1}) + \sum_{j < i} \beta_{k(j)} (x_{k(j)} - x_{k(j-1)}) \\
 =& f_k(x_{k1}) + \sum_{j < i} \sum_{j' \leq j} d_{k(j')} (x_{k(j)} - x_{k(j-1)})\\
 =& f_k(x_{k1}) + \sum_{j' < i} d_{k(j')} \sum_{i > j \geq j'} (x_{k(j)} - x_{k(j-1)}) \\
 =& f_k(x_{k1}) + \sum_{j' < i} d_{k(j')} (x_{k(i)} - x_{k(j')})
\end{align*}
We can write this more compactly in matrix notations.
\begin{equation*}
\begin{split}
\left[ \begin{array}{c}
f_k(x_{k1}) \\
... \\
f_k(x_{kn}) 
\end{array} \right] =
\left[ \begin{array}{ccc}
    |x_{k1} - x_{k(1)}|_+ & ... & |x_{k1} - x_{k(n-1)}|_+ \\
    ... & & \\
    |x_{kn} - x_{k(1)}|_+ & ... & |x_{kn} - x_{k(n-1)}|_+ 
\end{array} \right]
\left[ \begin{array}{c}
    d_{k(1)} \\
    ... \\
    d_{k(n-1)}
\end{array} \right] 
+ c_k \equiv \Delta_k d_k + c_k
\end{split}
\end{equation*}

Where $\Delta_k$ is a $n\times n-1$ matrix such that $\Delta_k(i,j) = |x_{ki} - x_{k(j)}|_+$, $d_k = (d_{k(1)} ,..., d_{k(n-1)})$, and $c_k = f_k(x_{k1})$.

Because $f_k$ has to be centered, $c_k = - \frac{1}{n} \mathbf{1}_n^\tran \Delta_k d_k$, therefore:
\[
\Delta_k d_k + c_k \mathbf{1}_n = 
   \Delta_k d_k - \frac{1}{n} \mathbf{1}_n \mathbf{1}_n^\tran \Delta_k d_k = 
   \bar{\Delta}_k d_k 
\]
where $\bar{\Delta}_k \equiv \Delta_k - \frac{1}{n} \mathbf{1}_n \mathbf{1}_n^\tran \Delta_k$ is $\Delta_k$ with the mean of each column subtracted.

@@

The decoupled concave postprocessing stage optimization is again similar. Suppose $\hat{d}_k$'s are the output of optimization~\ref{opt:alternate_opt}, define $\hat{r} = Y - \sum_{k=1}^p \bar{\Delta}_k \hat{d}_k$. 
\begin{align}
\textrm{for all}&\textrm{ $k$ such that $\hat{d}_k = 0$: }  \nonumber \\
  \min_{d_k} & 
      \frac{1}{2n} \Bigl \| \hat{r} - \bar{\Delta}_k d_k \Bigr \|_2^2
      + \lambda_n \| \bar{\Delta}_k d_k \|_\infty 
      \label{opt:alternate_opt_concave}\\
 \trm{s.t.} & d_{k(2)}, \ldots, d_{k(n-1)} \leq 0 \qquad \trm{(concavity)} \nonumber
\end{align}

\textbf{Note:} in sparsistency analysis, we assume that an upper bound $L$ to the Lipschitz smoothness of the true function $f_0$ is known and that we constrain our estimate $\hat{f}$ to obey the same Lipschitz condition, that is, each $\hat{f}_k$ must be $L$-Lipschitz. This Lipschitz constraint can be easily added to our optimization program. In optimization~\ref{opt:alternate_opt}, we can enforce the Lipschitz condition by adding two constraints: $d_{k(1)} \geq - L$ and $\sum_{j=2}^{n-1} d_{k(j)} \leq L$. We emphasize that we use the Lipschitz constraint only in our theoretical analysis; all of our experiments do not impose any Lipschitz condition.

>>>>

The sparse convex additive model optimization in (\ref{np}) is a quadratic program (QP) with
$O(np)$ variables and $O(np)$ constraints. 
Directly applying a QP solver for $\bds{h},
\bds{\beta}$ would be computationally expensive for relatively large
$n$ and $p$. However, notice that variables in different feature
dimensions are only coupled in the term
$(Y_{i}-\sum_{k=1}^{p}h_{ki})^{2}$. Hence, we can apply the block
coordinate descent method, where in each step we solve the following
QP subproblem for $\{\bds{h}_{k\cdot},\bds{\beta}_{k\cdot}\}$ with the
other variables fixed. We present the one-block optimization in matrix notation:
\begin{align}
\min_{ \bds{f}_k, \bds{\beta}_k, \gamma_k}& \frac{1}{2n} \| \bds{r}_{\mh k} - \bds{f}_k \|_2^2 
     + \lambda \gamma_k \label{opt:1d_compact} \\
 \textrm{s.t. } & P \bds{f}_k = \diag(P \bds{x}_k)  \bds{\beta}_k \nonumber \\
   & D \bds{\beta}_k \leq 0 ,
   \quad -\gamma_k \mathbf{1}_n \leq \bds{f}_k \leq \gamma_k \mathbf{1}_n,
   \quad \mathbf{1}_n^\tran \bds{f}_k = 0 \nonumber
\end{align}

@@

% \begin{equation}
% \label{eqn:opt_1d}
% \begin{split}
%        \min_{\bds{h}_{k\cdot},\bds{\beta}_{k\cdot},\gamma_{k}} &
%              \ \frac{1}{2n}\sum_{i=1}^{n}\Bigl((Y_{i}-\bar{Y}
%                 -\sum_{r\neq{k}}f_{ri})-f_{ki}\Bigr)^{2} 
%                       + \lambda\gamma_{k} \\
%         \textrm{such that} & \ f_{k(i+1)} = f_{k(i)} + \beta_{k(i)}(x_{k(i+1)}-x_{k(i)}),\\
%         &\ \beta_{k(i+1)} \geq \beta_{k(i)}, \ -\gamma_{k}\leq f_{ki}\leq\gamma_{k}\\
%         &\  \sum_{i=1}^{n}f_{ki}=0, \ (\forall i).
% \end{split}
% \end{equation}
The extra variable $\gamma_{k}$ is introduced to deal with the $\ell_{\infty}$ norm. This QP subproblem involves $O(n)$ variables, $O(n)$ constraints and a sparse structure, 
which can be solved efficiently using optimization packages (e.g., MOSEK: \verb+http://www.mosek.com/+).  We cycle through all feature dimensions ($k$) from $1$ to $p$ multiple times until convergence.
Empirically, we observe that the algorithm converges in only a few cycles. We also implemented an ADMM solver for (\ref{np}), but found that it is not as efficient as this QP solver.

After optimization, the function estimator for any input data $\bds{x}_j$ is, according to (\ref{hyper}),
\begin{equation}
\begin{split}
\nonumber
      f(\bds{x}_j) & = \sum_{k=1}^{p}f_k(x_{kj})+\mu 
= \sum_{k=1}^{p}\max_{i} \{f_{ki}+\beta_{ki}(x_{kj}-x_{ki})\} +
      \mu.
\end{split}
\end{equation}

The univariate concave function estimation is a straightforward modification of optimization~\ref{opt:1d_compact}. We need only modify one linear inequality constraint to enforce that the subgradients must be non-increasing: $\beta_{k(i+1)} \leq \beta_{k(i)}$.

@@

It is easy to verify that $\beta_{k(i)} = \sum_{j \leq i} d_{k(j)}$ and 
\begin{align*}
f_k(x_{k(i)}) = & f_k(x_{k(i-1)}) + \beta_{k(i-1)}(x_{k(i)} - x_{k(i-1)}) \\
 =& f_k(x_{k1}) + \sum_{j < i} \beta_{k(j)} (x_{k(j)} - x_{k(j-1)}) \\
 =& f_k(x_{k1}) + \sum_{j < i} \sum_{j' \leq j} d_{k(j')} (x_{k(j)} - x_{k(j-1)})\\
 =& f_k(x_{k1}) + \sum_{j' < i} d_{k(j')} \sum_{i > j \geq j'} (x_{k(j)} - x_{k(j-1)}) \\
 =& f_k(x_{k1}) + \sum_{j' < i} d_{k(j')} (x_{k(i)} - x_{k(j')})
\end{align*}
We can write this more compactly in matrix notations.
\begin{equation*}
\begin{split}
\left[ \begin{array}{c}
f_k(x_{k1}) \\
... \\
f_k(x_{kn}) 
\end{array} \right] =
\left[ \begin{array}{ccc}
    |x_{k1} - x_{k(1)}|_+ & ... & |x_{k1} - x_{k(n-1)}|_+ \\
    ... & & \\
    |x_{kn} - x_{k(1)}|_+ & ... & |x_{kn} - x_{k(n-1)}|_+ 
\end{array} \right]
\left[ \begin{array}{c}
    d_{k(1)} \\
    ... \\
    d_{k(n-1)}
\end{array} \right] 
+ \mu_k \equiv \Delta_k d_k + \mu_k
\end{split}
\end{equation*}

Where $\Delta_k$ is a $n\times n-1$ matrix such that $\Delta_k(i,j) = |x_{ki} - x_{k(j)}|_+$, $d_k = (d_{k(1)} ,..., d_{k(n-1)})$, and $\mu_k = f_k(x_{k1})$.

Because $f_k$ has to be centered, $\mu_k = - \frac{1}{n} \mathbf{1}_n^\tran \Delta_k d_k$, therefore:
\[
\Delta_k d_k + \mu_k \mathbf{1}_n = 
   \Delta_k d_k - \frac{1}{n} \mathbf{1}_n \mathbf{1}_n^\tran \Delta_k d_k = 
   \bar{\Delta}_k d_k 
\]
where $\bar{\Delta}_k \equiv \Delta_k - \frac{1}{n} \mathbf{1}_n \mathbf{1}_n^\tran \Delta_k$ is $\Delta_k$ with the mean of each column subtracted.

@@

The decoupled concave postprocessing stage optimization is again similar. Suppose $\hat{d}_k$'s are the output of optimization~\ref{opt:alternate_opt}, define $\hat{r} = Y - \sum_{k=1}^p \bar{\Delta}_k \hat{d}_k$. 
\begin{align}
\textrm{for all}&\textrm{ $k$ such that $\hat{d}_k = 0$: }  \nonumber \\
  \min_{c_k} & 
      \frac{1}{2n} \Bigl \| \hat{r} - \Delta_k c_k \Bigr \|_2^2
      + \lambda_n \| \Delta_k c_k \|_\infty 
      \label{opt:alternate_opt_concave}\\
 \trm{s.t.} & c_{k(2)}, \ldots, c_{k(n-1)} \leq 0 \qquad \trm{(concavity)} \nonumber
\end{align}

We can use either the off-centered $\Delta_k$ matrix or the centered $\bar{\Delta}_k$ matrix because the concave estimations are decoupled and hence suffer no identifiability problems.\\

\begin{remark}
in sparsistency analysis, we assume that $L$, an upper bound to the coordinate-wise Lipschitz smoothness of the true function $f_0$, is known and that we constrain our estimate $\hat{f}$ to obey the same Lipschitz condition, that is, each $\hat{f}_k$ must be $L$-Lipschitz. This Lipschitz constraint can be easily added to our optimization program. In optimization~\ref{opt:alternate_opt}, we can enforce the Lipschitz condition by adding two constraints: $d_{k(1)} \geq - L$ and $\sum_{j=2}^{n-1} d_{k(j)} \leq L$ (similarly for optimization~\ref{opt:alternate_opt_concave}). We emphasize that we use the Lipschitz constraint only in our theoretical analysis; all of our experiments do not impose any Lipschitz condition.
\end{remark}

### 2014-04-24 tex/scam/sparsistency.tex
experiment code

We divide our analysis into two parts. We first establish a sufficient
\emph{deterministic} condition for sparsistency.  We then consider the
stochastic setting and argue that the deterministic conditions hold with high probability.

@@

We follow \cite{Wain:09a} and define the \emph{restricted regression} purely for theoretical purposes.
\begin{definition}
In \emph{restricted regression}, we restrict the indices $k$ in
optimization (\ref{opt:alternate_opt}) to lie in the support $S$ instead of ranging from $1,...,p$. 
\end{definition}

@@

Then the following are true:
\begin{enumerate}
\item Let $\hat{d}_k = 0$ for $k \in S^c$, then \{$\hat{d}_k\}_{k=1,...,p}$ is an optimal solution to optimization~\ref{opt:alternate_opt}. Furthermore, any solution to the optimization program \ref{opt:alternate_opt} must be zero on $S^c$.
\item For all $k \in S^c$, the solution to optimization~\ref{opt:alternate_opt_concave} must be 0 and must be unique.
\end{enumerate}

>>>>

We divide our analysis into two parts. We first establish a sufficient deterministic condition for sparsistency.  We then consider the
stochastic setting and argue that the deterministic conditions hold with high probability.

@@

We construct an additive convex solution $\{\hat{d}_k\}_{k=1,...,p}$ that is zero for $k \in S^c$ and show that it satisfies the KKT conditions for optimality of optimization~\ref{opt:alternate_opt}. We define $\hat{d}_k$ for $k \in S$ to be a solution to the restricted regression (defined below). We will then also show that $\hat{c}_k = 0$ satisfies the optimality condition of optimization~\ref{opt:alternate_opt_concave} for all $k \in S^c$.

\begin{definition}
\label{def:restricted_regression}
In \emph{restricted regression}, we restrict the indices $k$ in
optimization (\ref{opt:alternate_opt}) to lie in the set $S$ instead of ranging from $1,...,p$. 
\end{definition}

@@

Then the following are true:
\begin{enumerate}
\item Let $\hat{d}_k = 0$ for $k \in S^c$, then \{$\hat{d}_k\}_{k=1,...,p}$ is an optimal solution to optimization~\ref{opt:alternate_opt}. Furthermore, any solution to the optimization program \ref{opt:alternate_opt} must be zero on $S^c$.
\item For all $k \in S^c$, the solution $\hat{c}_k$ to optimization~\ref{opt:alternate_opt_concave} must be 0 and must be unique.
\end{enumerate}
